Structural replication of nanoporous media using procedural noise

### 2017-02-06 ms.tex
Renskrevet mengder, endel nytt

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract} 
In this paper, we aim to develop a method for producing statistically correct porous media using noise functions. In short, the proposed model is a binary function that returns either true if the input point is inside pore space, or false if within a wall. By using procedural noise, we are not restricted to generating media within a limited box, as the noise functions is continuous on all scales. It is therefore possible to generate geometries at arbitrary scales. In addition, this method is extremely fast and has low memory footprints, requiring very little computational power. After presenting such a model, we choose a method to measure the structural properties of nanoporous media that are created either through regular methods (e.g. molecular dynamics) or with our method. The radial distribution function $g(r)$ is chosen as a measure of the structural properties to compare the simulated geometry to our model. Using this measure, we define a full likelihood framework for estimating model parameters given any data set, and continue by validating our own framework by analysing procedural noise data sets with known input parameters. Finally, we generate a nanoporous media with an expand-and-quench technique using molecular dynamics and show that our model correctly reproduces several of the internal properties of the data set, including $g(r)$, porosity and surface area.  \todo{Abstract boer tenkes noe gjennom: faa frem hovedpoenger.}

@@

%During recent years, increase in computational power has allowed for a vast improvement  
\todo{Skriv en fin intro! Ogsaa om nanoporous media}
Properties of structural patterns that regularly occur in nature are often investigated either through direct observations or through simulations of theoretical models. Model simulations typically require a large computational framework to accurately reproduce the observed effects, while experimental data can be both expensive and hard to obtain, in addition to being riddled with artefacts and instrumental errors.

Procedural generation of physical structures have several advantages over conventional simulations. For example, when producing simulations of nanoporous silica, we are required to calculate and propagate the physical properties of each particle for every time-step throughout the simulation, using a computational framework such as LAMMPS \cite{plimpton1995fast}. This is both time-consuming, and the computational cost can sometimes be be quite large. In addition, the simulation is constrained to a box, so particles are not allowed to move outside the defined boundaries. On the other hand, using procedural noise to carve out nanoporous media is both extremely fast and computationally cost-efficient, in addition to being continuous on all scales. This means that it is possible to simulate 
\todo{Nevn finite size effects}
%In addition, by creating a model that yields properties that are statistically consistent with observations, we are ar

In this paper, we aim to investigate whether procedural methods are suitable for producing statistically correct 3D physical structures, focusing on nanoporous media. In a sense, we are combining well-known methods from computer games and real-time graphics-oriented software with experimental observations and simulations. Procedural methods for automated generation of computer game content have been around since the early 1980's, and in those early days algorithms were less advanced, often yielding plain, simple geometrical structures. Nowadays, almost every single computer game produced will contain some level of automated generated content, from textures to level design to landscapes and gameplay.

We do not claim that procedural noise methods completely reproduces all structural properties of nanoporous silica, but show that our approach can be used to efficiently create materials that are structurally similar to simulations. Naturally, this all depends on the statistics used for comparing the two models, which in this case is based on the radial distribution function $g(r)$ (RDF). 
\todo{Nevne at kjemi naturligvis ikke vil være riktig siden antall naboer per atom ikke er riktig. }
%\section{Method}
\section{Procedural content generation}
By procedural content generation, we generally mean a mathematical algorithm that uses limited input data to create complex, sometimes infinite repeating patterns. A firm example would be fractals, such as the Julaia/Mandelbrot [ref mandelbrot] set, where one tests whether the complex quadratic recurrence equation $z_{n+1} = z_n^2 +C$ for $z \in \mathbb C$ converges for each position in a 2D complex plane. With this seemingly simple equation, the level of complexity generated on the boundary between convergence and divergence is infinite, with self-similar patterns repeating themselves to arbitrary detail.
\begin{figure}
\includegraphics[width=.5\textwidth]{orphancity.png}
\caption{Example of city building using procedural methods: Landscape, city infrastructure, buildings and all textures are all automatically generated using procedural methods. Created by one of the authors. \todo{Denne er jo med bare for moro skyld, boer jo fjernes} }
\label{fig:orphancity}
\end{figure}

Another example, widely used in contemporary game content generation, is having prefabs (pre-fabricated 3D models such as doors, walls, tiles and roof segments) automatically being stitched together to create buildings. When designing a city, instead of creating each building manually, you only need specify the position, area and height of a structure. Then, by using an algorithm that combines these prefabs to render the building, it is possible to create cities of arbitrary size with a single command. Naturally, it is also possible to create procedural methods that design the layout and placement of these buildings. See figure \ref{fig:orphancity} for an example of such a city generator tool, developed by one of the authors of this paper.

A much simpler example would be generating a 1 (or 2)-dimensional landscape by choosing random points on a grid and performing a spline interpolation between them to create the illusion of a smooth landscape. In fact, this is the very essence of procedural noise, which we are focusing on in this paper, and the first person to create such a fast algorithm for N-dimensional spline interpolation between pseudo-random points was Ken Perlin.

@@

\begin{figure}
\includegraphics[width=.5\textwidth]{noise_comparison.png}
%\mbox{\epsfig{file=perlinstats.eps, width=\linewidth}}
\caption{Normalized distributions of Perlin noise and simplex noise compared with a
  Gaussian distribution. Note how both noise functions produce
  near-Gaussian, but slightly different distributions.}
\label{fig:properties}
\end{figure}
Standard Perlin noise is known for having several anisotropic features
\citep{lagae:2010}. Perlin noise also gives rise to a ``textile''
pattern in the $x$ and $y$-directions, which is evident from the upper
left part of figure \ref{fig:noise}, depicting 2D Perlin noise in
a single octave. The upper-right part shows the same node for simplex noise,
with less obvious patterns. However, when combining several octaves and adding a random shift to the coordinates,
as depicted in the two lower parts, these anisotropic patterns tend to cancel
out.  In addition, we have calculated the statistical distributions for Perlin and simplex noise
with $\sigma = 0.3$, and have plotted the results in figure \ref{fig:properties}.
Each procedural noise method has distributions that are close to normal, 
but with some asymmetrical properties.

\subsection{Multiple octaves: creating patterns}
\label{sec:octaves}
The different octaves from procedural noise functions can be combined
to produce a wide range of nature-like patterns. A generic
N-dimensional procedural function $\Phi(\vec x): \mathbb R^N \to
\mathbb R$ could be expressed as
\begin{equation}
\label{eq:procedural}
 \Phi(\vec x) = \Theta \Big(\sum_k^\omega \kappa (k) P(  f_s (\vec x,k) )) \Big),
 \end{equation}
where $f_s(\vec x,k)$ describes the frequency scale modifier (e.g. $f_s(\vec x,k) =
k\vec x$), $\omega$ is the amount of octaves (modes), $\kappa(k)$ is the octave amplitude (e.g. $\kappa(k) = 1/k$)
and $\Theta(x)$ an overall modifier (e.g. $\Theta(x) = x$ or
$\Theta(x) = \frac{1}{x}$). In a sense, $\kappa(k)$ is similar to Fourier coefficients. 
In its simplest case, summing directly over the octaves with amplitude $1/k$ results in 
\begin{equation}
\label{eq:perlinlinear}
 \Phi(\vec x) = \sum_k^\omega \frac{1}{k} P( 2k\vec x).
\end{equation}
In 1 dimension, equation \ref{eq:perlinlinear} with 1, 3 and 30 octaves is depicted in figure \ref{fig:1dperlin}, while the 2D procedural noise pattern depicted in the lower
part of figure \ref{fig:noise}. Finally, since simplex noise both scales better with higher dimensions and produces less asymmetric effects, we choose to use this noise method over regular Perlin noise in our method.

@@

\section{Model}
In this paper, we are interested in developing a model using simplex noise to reproduce nanoporous structures that are as statistically close to actual simulations of porous media. For a given data set and a model, we wish to estimate the parameters that best represent the data. This is performed through a full likelihood analysis, where we in addition validate the model by comparing the radial distribution function $g(r)$, porosity and surface area of the system. The simulated system is confined to a cube of size $\SI{159} {\angstrom}^3$, but since the procedural noise function is continuous on all scales, the box could be of arbitrary size.

\begin{figure*}
\includegraphics[width=.95\textwidth]{model_examples.png}
\caption{Examples of 3D simplex noise structures in nanoporous media. Left image: low persistence (large scale dominates), middle image: medium persistence (small scales starts dominating) and right image: high threshold (more atoms are discarded). }
\label{fig:model_example}
\end{figure*}
We construct a generic sum over procedural noise octaves with the expression: 
\begin{equation}
  M_1(\bar \theta) = M_1(\vec x, \psi, \Omega, \nu) = \sum_{\omega=0}^{\Omega} \psi^{\omega -1}   N\big((\vec x + \sigma \cdot \omega)\nu \cdot 2^{\omega-1} \big),
\label{eq:noisemodel1}
\end{equation}
where $\Omega$ is the total amount of octaves, $\psi$ the persistence (scale falloff), $N$ a generic $N$-dimensional noise function, $\sigma$ a predefined random vector that removes anisotropic artefacts and $\nu$ the frequency modifier representing the initial largest scales in the model. For simplicity, $\bar \theta$ here denotes the set of model parameters. For each mode, the frequency is thus multiplied, while the overall amplitude is damped by a given persistence. In the end, we simulate a large set of bulk particles without any walls, and use the noise function to remove particles that are not in a procedural void given by a threshold $\tau$. Three examples of this model are depicted in figure \ref{fig:model_example}, where we have used $\Omega=6$ octaves and the scale is $\nu=0.01$ and cut-off $\tau=0$. The left image depicts low persistence ($\psi = 0.0$), where only the first octave dominates (corresponding to large scales). The middle picture shows the same parameters with a higher persistence $(\psi = 0.5)$, where smaller scales become more dominating. In the rightmost picture, the effect of the cut-off threshold $\tau = -0.3$ is shown.

\subsection{Model Measure}
In order to perform data analysis on the model parameters, we developed a code that uses Monte Carlo Markov chains in order to produce samples for building the likelihood. Before doing so, we need a model measure - a quantitative method that characterizes the internal structure of the porous media. This quantity should be computed for both the original data set and the one produced by our noise model.

Ideally, the model measure should pick up all geometrical properties that are relevant for the physical study of interest. For instance, transport in porous media heavily depends on the pore size distribution and porosity, whereas chemical reactons are sensitive to surface area. \todo{referanser, malthe?} In this paper, we choose to focus on the radial distribution function $g(r)$ as defined and implemented in LAMMPS \cite{plimpton1995fast}. $g(r)$ describes how density varies as a function of distance from a reference particle, picking up both the intramolecular and intermolecular structure. We have chosen to ignore the short range contributions of $g(r)$ since the pore geometry does not depend on the intramolecular structure of SiO$_2$, but only the density variations at larger distances. Hereafter, $g_d(r)$ denotes the RDF calculated from simulated or experimental data, while $g_{M(\bar \theta)}(r)$ represents the RDF based on the procedural noise model $M(\bar \theta)$, where $\bar \theta$ are the set of model parameters.

The method used for calculating porosity of a media is defined as in \cite{gelb1998characterization}, and we obtain the surface area of a porous media by using marching cubes triangulation to sum the surface area of the generated triangles. As discussed in \cite{gelb1998characterization}, the definition of surface area is ill-defined in atomic simulations, because a definition of pore space is needed. In the marching cubes triangulation, we generate a voxelized grid, the scalar field, with values being the distance from the voxel center to the nearest atom. The isovalue threshold defining the surface is chosen to be \SI{2}{\angstrom}, large nough to have a dense pore matrix, but small enough to have most of the available pore space defined as such. The important note here is that the same definition of surface is being used in all compared systems.

@@

The posterior probability is defined as the probability of the parameters $\bar \theta$ given the evidence $d$, or $P(\bar \theta | d)$. Bayes' theorem states that 
\begin{equation}
P(\bar \theta | d) = \frac{P(d | \bar \theta)P(\bar \theta)}{P(d)} \propto P(d | \bar \theta)P(\bar \theta),
\end{equation}
where $P(d | \bar \theta)$ is called the Likelihood $\mathcal L$ (the probability of evidence given parameters) and $P(\bar \theta)$ is the prior (the beliefs about the quantity before some evidence is taken into account). $P(d)$ is called the evidence, and is the probability of the observations (here assumed to be 1). Since the prior is a constant, we can therefore assume that the posterior distribution is proportional to the likelihood, or $P(d | \bar \theta) \propto \mathcal L$. In other words, to calculate the posterior and obtaining the best fit model parameters, we need only to estimate the likelihood function, which is connected to the $\chi^2$ by 
\begin{equation}
\mathcal L \propto e^{-\chi^2}.
\end{equation}

In order to test whether two 3D data sets are statistically equivalent up to the model measure $g(r)$, we calculate the $\chi^2$ between the two one-dimensional RDFs. Pearson's $\chi^2$ is defined as 
\begin{equation}
  \chi^2 = \sum_r \frac{ \Big(g_d(r) - g_{M(\bar \theta)}(r) \Big)^2}{g_{M(\bar \theta)}(r)^2},
\end{equation}
so for low values of $\chi^2$, the better the fit between data and model (up to $g(r)$). For a single-parameter model, we could simply just iterate over the lone parameter to obtain the minimum $\chi^2$ (where the best-fit parameter value is), however, with increased number of parameters comes increased computational cost. If we divide the parameter space into $N$ bins, the number of $\chi^2$ computations scales as $N^d$, where $d$ is the number of model parameters. To be able to build the full $\chi^2$ (and likelihood) parameter landscape, we there need to use Monte Carlo Markov chain methods (MCMC).

We are now interested in maximising the likelihood $\mathcal L$ as opposed to minimising the $\chi^2$, with the goal of mapping out the full $d$-dimensional likelihood space for the model parameters. We build the $d$-dimensional likelihood by letting random walkers traverse the parameter space guided by the MCMC test: for each step $M_n(\bar \theta)$ in parameter space, we calculate the likelihood for this parameter configuration. Then, we create a new set of parameters for the next step $M_{n+1}(\bar \theta)$ by choosing new parameters randomly from a $d$-dimensional sphere around the previous parameter composition. The model also selects a new random seed. We calculate the likelihood at the new step and perform the MCMC test: accept the new step if $\frac{\mathcal L_n}{\mathcal L_n+1} \ge 1$, or reject if the value is lower than a uniform random value $\in [0,1]$.

@@

\section{Results}
We now turn our attention to applying the MCMC framework on two kinds of data. First, we wish to test and verify the code by performing a full data analysis on mock data with known input parameters. This enables us to investigate the posterior distribution, and determine whether any potential problems could arise with the likelihood landscape, such as radom walkers trapped in local minima. We then continue with a full analysis of simulated SiO$_2$ data, and see whether the resulting best-fit parameter data set is consistent with the SiO$_2$ data set.

\subsection{Model verification: Parameter estimation of mock data}
\begin{figure}[htb!]
\includegraphics[width=.45\textwidth]{model_test.png}
\caption{Simulated mock data used for the initial model testing analysis. Input parameters are $\nu=0.03$,$\tau=-0.02$ and $\psi=0.6$.}
\label{fig:mockdata}
\end{figure}
In order to verify our likelihood code, we perform a model verification test on a mock data set with known input values. In this example, we only use standard simplex noise with three free parameters: the persistence $\psi$, threshold $\tau$ and scale $\nu$. We simulate a data set of $\SI{159} {\angstrom}^3$ of $\sim 255 000$ atoms, and perform the cut-off given input parameters $\Omega=3$, $\Psi = 0.6$, $\tau=-0.2$, $\nu=0.03$. The resulting 3D structure is depicted in figure \ref{fig:mockdata}.

We continue by performing a full MCMC analysis of the mock data with using a parallelized code running on a 48 core cluster. In order to produce 300 000 samples, about 48 hours was needed. The results from this analysis can be seen in figure \ref{fig:mockdataresults_pts}, where the upper panel depicts the marginalised 1D posteriors while the lower shows the marginalised 2D posteriors. The centre and left image shows that the input parameters of $\tau=-0.2$(threshold), $\psi=0.6$(persistence) and $\nu=0.03$ are all reproduced well within $1 \sigma$.

Note that there seems to be no problems with the posterior distribution in terms of local minima or multiple maxima.

\begin{figure*}
\includegraphics[width=.99\textwidth]{mock_data_results_pts.png}
\caption{Results from the model validation simulation. The upper panel displays the marginalised 1D posteriors for each parameter, while the lower panel depicts the marginalised 2D posteriors. Note that all input parameters were reproduced well within $1\sigma$ (inner contour line).}
\label{fig:mockdataresults_pts}
\end{figure*}

@@

\label{fig:porous_results1}
\end{figure*}

We now perform a parameter fitting analysis on realistic simulated nanoporous media. Using LAMMPS \cite{plimpton1995fast} and the Vashishta-potential \cite{vashishta1990interaction}, we simulate system consisting of $14^3$ betacristobalite unit cells of SiO$_2$, corresponding to system size of approximately \SI{100}{\angstrom}. The temperature is then increased to \SI{4500}{\kelvin} using NVT integration. After an initial period, we expand the box to \SI{159}{\angstrom} to achieve a porosity $\phi=0.75$. Then, the system is quenched by setting the temperature to \SI{300}{\kelvin}, resulting in the nanoporous media on the right side in figure \ref{fig:porous_vs_model}. By choosing a different seed for the initial velocities, it is easy to produce other samples.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract} 
In this paper, we aim to develop a method for producing statistically correct porous media using procedural noise functions. In short, the proposed model is a binary function that returns either true if the input point is inside pore space, or false if within a wall. By using procedural noise, we are not restricted to generating media within a limited box, as the noise functions is continuous on all scales. It is therefore possible to generate fractal-like geometries at arbitrary scales, circumventing the need for any physics simulations. In addition, this method is extremely fast and has low memory footprints, requiring very little computational power. After presenting such a model, we choose a method to measure the structural properties of nanoporous media that are created either through regular methods (e.g. molecular dynamics) or with our method. The radial distribution function $g(r)$ is chosen as a measure of the structural properties to compare the simulated geometry to our model. Using this measure, we define a full likelihood framework for estimating model parameters given any data set, and continue by validating our own framework by analysing procedural noise data sets with known input parameters. Finally, we generate a nanoporous media with an expand-and-quench technique using molecular dynamics and show that our model correctly reproduces several of the internal properties of the data set, including $g(r)$, porosity and surface area. 
  \todo{Abstract boer tenkes noe gjennom: faa frem hovedpoenger.}

@@

%During recent years, increase in computational power has allowed for a vast improvement  
\todo{Skriv en fin intro! Ogsaa om nanoporous media}
Properties of structural patterns that regularly occur in nature are often investigated either through direct observations or through simulations of theoretical models. Model simulations typically require a large computational framework to accurately reproduce the observed effects, while experimental data can be both expensive and hard to obtain, in addition to being riddled with artefacts and instrumental errors. While physical simulations yields statistically correct results, it is often hard to re-create features that are noticeable in experimental data, and therefore fit model parameters with observations. In this paper, we propose a new method for producing realistic-looking nanoporous media using only random noise fields, without the need for physics simulations. We do not claim that our method produces results that are more physically accurate than full model simulations, quite the contrary, we aim to show that it is possible to create fast, near-accurate and arbitrary large data sets of nanoporous media that are statistically similar to physics-based simulations with only the use of random noise fields.

Procedural generation of physical structures have several advantages over conventional simulations. For example, when producing simulations of nanoporous silica, we are required to calculate and propagate the physical properties of each particle for every time-step throughout the simulation, using a computational framework such as LAMMPS \cite{plimpton1995fast}. This is both time-consuming, and the computational cost can sometimes be be quite large. In addition, the simulation is constrained to a box, so particles are not allowed to move outside the defined boundaries. On the other hand, using procedural noise to carve out nanoporous media is both extremely fast and computationally cost-efficient, in addition to being continuous on all scales.

\todo{Nevn finite size effects}
%In addition, by creating a model that yields properties that are statistically consistent with observations, we are ar

Since we aim to investigate whether procedural methods are suitable for producing statistically correct 3D physical structures, we choose to focus our attention on silica-based nanoporous media. In a sense, we are combining well-known methods from computer games and real-time graphics-oriented software with experimental observations and simulations. Procedural methods for automated generation of computer game content have been around since the early 1980's, and in those early days algorithms were less advanced, often yielding plain, simple geometrical structures. Nowadays, almost every single computer game produced will contain some level of automated generated content, from textures to level design to landscapes and gameplay.

We do not claim that procedural noise methods completely reproduces all structural properties of nanoporous silica, but show that our approach can be used to efficiently create materials that are structurally similar to simulations. Naturally, this all depends on the statistics used for comparing the two models, which in this case is based on the radial distribution function $g(r)$ (RDF). 
\todo{Nevne at kjemi naturligvis ikke vil være riktig siden antall naboer per atom ikke er riktig. }
%\section{Method}
\section{Procedural content generation}
By procedural content generation, we generally mean a mathematical algorithm that uses limited input data to create complex, sometimes infinite repeating patterns. A firm example would be fractals, such as the Julia/Mandelbrot [ref Mandelbrot] set, where one tests whether the complex quadratic recurrence equation $z_{n+1} = z_n^2 +C$ for $z \in \mathbb C$ converges for each position in a 2D complex plane. With this seemingly simple equation, the level of complexity generated on the boundary between convergence and divergence is infinite, with self-similar patterns repeating themselves to arbitrary detail.
\begin{figure}
\includegraphics[width=.5\textwidth]{orphancity.png}
\caption{Example of city building using procedural methods: Landscape, city infrastructure, buildings and all textures are all automatically generated using procedural methods. Created by one of the authors. \todo{Denne er jo med bare for moro skyld, boer jo fjernes} }
\label{fig:orphancity}
\end{figure}

Another example, widely used in contemporary game content generation, is having prefabs, or pre-fabricated 3D models such as doors, walls, tiles and roof segments automatically being stitched together to create buildings. When designing a city, instead of creating each building manually, you only need specify the position, area and height of a structure. Then, by using an algorithm that combines these prefabs to render the building, it is possible to create cities of arbitrary size with a single command. Naturally, it is also possible to create procedural methods that design the layout and placement of these buildings. See figure \ref{fig:orphancity} for an example of such a city generator tool, developed by one of the authors of this paper.

A much simpler example would be generating a 1 (or 2)-dimensional terrain by choosing random points on a grid and performing a spline interpolation between them to create the illusion of a smooth landscape. In fact, this is the very essence of procedural noise, which we are focusing on in our model, and the first person to create such a fast algorithm for N-dimensional spline interpolation between pseudo-random points was Ken Perlin.

@@

\begin{figure}
\includegraphics[width=.5\textwidth]{noise_comparison.png}
%\mbox{\epsfig{file=perlinstats.eps, width=\linewidth}}
\caption{Normalized distributions of Perlin noise and simplex noise compared with a
  Gaussian distribution. Note how both noise functions produce
  near-Gaussian, but slightly different distributions.}
\label{fig:properties}
\end{figure}
Standard Perlin noise is known for having several anisotropic features
\citep{lagae:2010}. Perlin noise also gives rise to a ``textile''
pattern in the $x$ and $y$-directions, which is evident from the upper
left part of figure \ref{fig:noise}, depicting 2D Perlin noise in
a single octave. The upper-right part shows the same node for simplex noise,
with less obvious ``textile'' features. However, when combining several octaves and adding a random shift to the coordinates, as depicted in the two lower parts, these anisotropic patterns tend to cancel
out.  In addition, we have calculated the statistical distributions for Perlin and simplex noise
with $\sigma = 0.35$, and have plotted the results in figure \ref{fig:properties}.
Each procedural noise method has distributions that are close to normal, 
but with some asymmetrical properties.

\subsection{Multiple octaves: creating patterns}
\label{sec:octaves}
The different octaves from procedural noise functions can be combined
to produce a wide range of nature-like patterns. A generic
N-dimensional procedural function $\Phi(\vec x): \mathbb R^N \to
\mathbb R$ could be expressed as
\begin{equation}
\label{eq:procedural}
 \Phi(\vec x) = \Theta \Big(\sum_k^\Omega \kappa (k) P(  f_s (\vec x,k) )) \Big),
 \end{equation}
where $f_s(\vec x,k)$ describes the frequency scale modifier (e.g. $f_s(\vec x,k) =
k\vec x$), $\omega$ is the amount of octaves (modes), $\kappa(k)$ is the octave amplitude (e.g. $\kappa(k) = 1/k$)
and $\Theta(x)$ an overall modifier (e.g. $\Theta(x) = x$ or
$\Theta(x) = \frac{1}{x}$). In a sense, $\kappa(k)$ is similar to Fourier coefficients. 
In its simplest case, summing directly over the octaves with amplitude $1/k$ results in 
\begin{equation}
\label{eq:perlinlinear}
 \Phi(\vec x) = \sum_k^\Omega \frac{1}{k} P( 2k\vec x).
\end{equation}
In one dimension, we show equation \ref{eq:perlinlinear} with 1, 3 and 30 octaves in figure \ref{fig:1dperlin}, while the 2D procedural noise pattern depicted in the lower
part of figure \ref{fig:noise}. Finally, since simplex noise both scales better with higher dimensions and produces less asymmetric effects, we choose to use simplex over regular Perlin noise as the main noise function throughout this paper.

@@

\section{Model}
In this paper, we are interested in developing a model using simplex noise to reproduce nanoporous structures that are as statistically close to physical simulations of porous media. For a given data set and a model, we wish to estimate the parameters that best represent the data. This is performed through a full likelihood analysis in order to obtain the posterior probability distribution of the parameters. The model measure of the analysis is the radial distribution function $g(r)$, for which the parameters will be maximised to fit. In addition, we also investigate the porosity and surface area of the system. The simulated system is confined to a cube of size $\SI{159} {\angstrom}^3$, but since the procedural noise function is continuous on all scales, the box could be of arbitrary size.

\begin{figure*}
\includegraphics[width=.95\textwidth]{model_examples.png}
\caption{Examples of 3D simplex noise structures in nanoporous media. Left image: low persistence (large scale dominates), middle image: medium persistence (small scales starts dominating) and right image: high threshold (more atoms are discarded). }
\label{fig:model_example}
\end{figure*}
We construct a generic sum over procedural noise octaves with the expression: 
\begin{equation}
  M_1(\bar \theta) = M_1(\vec x, \psi, \Omega, \nu) = \sum_{\omega=0}^{\Omega} \psi^{\omega -1}   N\big((\vec x + \sigma \cdot \omega)\nu \cdot 2^{\omega-1} \big),
\label{eq:noisemodel1}
\end{equation}
where $\Omega$ is the total amount of octaves, $\psi$ the persistence (scale falloff), $N$ a generic $N$-dimensional noise function, $\sigma$ a predefined random vector that removes anisotropic artefacts and $\nu$ the frequency modifier representing the initial largest scales in the model. For simplicity, $\bar \theta$ here denotes the set of model parameters. For each mode, the frequency is thus multiplied, while the overall amplitude is damped by a given persistence. In the end, we simulate a large set of bulk particles without any walls, and use the noise function to remove particles that are not within a procedural void given by a threshold $\tau$. Three examples of this model are depicted in figure \ref{fig:model_example}, where we have used $\Omega=6$ octaves and the scale is $\nu=0.01$ and cut-off $\tau=0$. The left image depicts low persistence ($\psi = 0.0$), where only the first octave dominates (corresponding to large scales). The middle picture shows the same parameters with a higher persistence $(\psi = 0.5)$, where smaller scales become more dominating. In the rightmost picture, the effect of the cut-off threshold $\tau = -0.3$ is shown to emphasize threshold cut-off.

\subsection{Model Measure}
In order to perform data analysis on the model parameters, we developed a code that uses Monte Carlo Markov chains to produce samples for building the posterior probability distribution for the parameters. Before doing so, we need a model measure - a quantitative method that characterizes the internal structure of the porous media. This quantity should be computed for both the original data set and the one produced by our noise model.

Ideally, the model measure should pick up all geometrical properties that are relevant for the physical study of interest. For instance, transport in porous media heavily depends on the pore size distribution and porosity, whereas chemical reactons are sensitive to surface area. \todo{referanser, malthe?} In this paper, we choose to focus on the radial distribution function $g(r)$ as defined and implemented in LAMMPS \cite{plimpton1995fast}. $g(r)$ describes how density varies as a function of distance from a reference particle, picking up both the intramolecular and intermolecular structure. We have chosen to ignore the short range contributions of $g(r)$ since the pore geometry does not depend on the intra-molecular structure of SiO$_2$, but only the density variations at larger distances. Hereafter, $g_d(r)$ denotes the RDF calculated from simulated or experimental data, while $g_{M(\bar \theta)}(r)$ represents the RDF based on the procedural noise model $M(\bar \theta)$, where $\bar \theta$ is the set of model parameters.

The method used for calculating porosity of a media is defined as in \cite{gelb1998characterization}, and we obtain the surface area of a porous media by using marching cubes triangulation to sum the surface area of the generated triangles. As discussed in \cite{gelb1998characterization}, the definition of surface area is ill-defined in atomic simulations, because a clear definition of pore space is needed. In the marching cubes triangulation, we create a voxelized grid, a scalar field with values being the distance from the voxel centre to the nearest atom. The isovalue threshold defining the surface is chosen to be \SI{2}{\angstrom}, large enough to contribute to a dense pore matrix, but small enough to have most of the available pore space defined as such. The important note here is that the same definition of both surface and porosity is being used in all systems that are compared, physical simulations, procedural methods or data.

@@

The posterior probability distribution is defined as the probability of the parameters $\bar \theta$ given the evidence $d$, or $P(\bar \theta | d)$. Bayes' theorem states that 
\begin{equation}
P(\bar \theta | d) = \frac{P(d | \bar \theta)P(\bar \theta)}{P(d)} \propto P(d | \bar \theta)P(\bar \theta),
\end{equation}
where $P(d | \bar \theta)$ is called the Likelihood $\mathcal L$ (the probability of evidence given parameters) and $P(\bar \theta)$ is the prior (the beliefs about the quantity before some evidence is taken into account). $P(d)$ is called the evidence, and is the probability of the observations (here assumed to be 1). Prior parameters are restricted to freely roam within a minimum and maximum value, so we can therefore assume that the posterior distribution is proportional to the likelihood, or $P(d | \bar \theta) \propto \mathcal L$. In other words, to calculate the posterior probability distribution and obtaining the best fit model parameters, we need only to estimate the likelihood function, which is connected to the $\chi^2$ by 
\begin{equation}
\mathcal L \propto e^{-\chi^2}.
\end{equation}

In order to estimate whether two 3D data sets are statistically equivalent up to the model measure $g(r)$, we calculate the $\chi^2$ between the two one-dimensional RDFs. Pearson's $\chi^2$ is defined as 
\begin{equation}
  \chi^2 = \sum_r \frac{ \Big(g_d(r) - g_{M(\bar \theta)}(r) \Big)^2}{g_{M(\bar \theta)}(r)^2},
\end{equation}
so for low values of $\chi^2$, the better the fit between data and model, up to $g(r)$. For a single-parameter model, we could simply just iterate over the lone parameter to obtain the minimum $\chi^2$ (where the best-fit parameter value resides), however, with increased number of parameters comes increased computational cost. If we divide the parameter space into $N$ bins, the number of $\chi^2$ computations scales as $N^d$, where $d$ is the number of model parameters. To be able to build the full landscape of the posterior probability distribution, we need to use the Monte Carlo Markov chain (MCMC) method .

We are now interested in maximising the likelihood $\mathcal L$ as opposed to minimising the $\chi^2$, with the goal of mapping out the full $d$-dimensional likelihood space for the model parameters. We build the $d$-dimensional likelihood by letting random walkers traverse the parameter space guided by the MCMC test: for each step $M(\bar \theta_{n})$ in parameter space, we calculate the likelihood $\mathcal L_n$ for this particular parameter configuration. Then, for the next step, we create a new set of parameters $M(\bar \theta_{n+1})$ by choosing new parameters randomly from a $d$-dimensional sphere around the previous parameter composition. The model also selects a new random seed. We calculate the proposed likelihood $\mathcal L_{n+1}$ at this new step and perform the MCMC test: accept the new step if $\frac{\mathcal L_n}{\mathcal L_n+1} \ge 1$, or reject if the value is lower than a uniform random value $\in [0,1]$.

@@

\section{Results}
We now turn our attention to applying the MCMC framework on two kinds of data. First, we wish to test and verify the code by performing a full data analysis on mock data with known input parameters. This enables us to investigate the posterior distribution, and determine whether any potential problems could arise with the likelihood landscape, such as random walkers trapped in local minima. In addition, we will obtain an intuition on the width of the estimated parameter distributions. But most importantly, we will have verified that our likelihood code is working correctly if we successfully reproduce the original input parameters. We then continue with a full analysis of simulated SiO$_2$ data, and see whether the resulting best-fit parameter data set is consistent with the SiO$_2$ data set.

\subsection{Model verification: Parameter estimation of mock data}
\begin{figure}[htb!]
\includegraphics[width=.45\textwidth]{model_test.png}
\caption{Simulated mock data used for the initial model testing analysis. Input parameters are $\nu=0.03$,$\tau=-0.02$ and $\psi=0.6$.}
\label{fig:mockdata}
\end{figure}
In order to verify our likelihood code, we perform a model verification test on a mock data set with known input values based on equation \ref{eq:noisemodel1}. In this example, we only use standard simplex noise with three free parameters: the persistence $\psi$, threshold $\tau$ and scale $\nu$. We simulate a data set of size $\SI{159} {\angstrom}^3$ with $\sim 255 000$ atoms, and perform the cut-off given by input parameters $\Omega=3$, $\Psi = 0.6$, $\tau=-0.2$ and $\nu=0.03$. The resulting 3D structure is depicted in figure \ref{fig:mockdata}.

We continue by executing a full MCMC analysis on the mock data by using a parallelized code written in C++ running on a 48 core cluster. About 48 hours was needed to produce about 300 000 samples. The results from this analysis can be seen in figure \ref{fig:mockdataresults_pts}, where the upper panel depicts the marginalised 1D posteriors while the lower shows the marginalised 2D posteriors. The centre and left image shows that the input parameters of $\tau=-0.2$(threshold), $\psi=0.6$(persistence) and $\nu=0.03$ are all reproduced well within $1 \sigma$. Note that there seems to be no problems with the posterior distribution in terms of local minima or multiple maxima. In addition, the results show that all the parameters are correlated with $r \sim 0.85$, while the width of the marginalised distributions vary by $\sigma_\tau = 0.02 \sim 10\%$, $\sigma_\Psi = 0.062 \sim 20\%$ and $\sigma_\nu = 0.01 \sim 33\%$.

\begin{figure*}
\includegraphics[width=.99\textwidth]{mock_data_results_pts.png}
\caption{Results from the model validation simulation. The upper panel displays the marginalised 1D posteriors for each parameter, while the lower panel depicts the marginalised 2D posteriors. Note that all input parameters were reproduced well within $1 \sigma$ (inner contour line).}
\label{fig:mockdataresults_pts}
\end{figure*}

@@

\label{fig:porous_results1}
\end{figure*}
Having verified that the likelihood code yields correct results when analysing mock data, we turn to 
performing a parameter fitting analysis on realistic simulated nanoporous media. Using LAMMPS \cite{plimpton1995fast} and the Vashishta-potential \cite{vashishta1990interaction}, we simulate system consisting of $14^3$ betacristobalite unit cells of SiO$_2$, corresponding to system size of approximately \SI{100}{\angstrom}. The temperature is then increased to \SI{4500}{\kelvin} using NVT integration. After an initial period, we expand the box to \SI{159}{\angstrom} to achieve a porosity $\phi=0.75$. Then, the system is quenched by setting the temperature to \SI{300}{\kelvin}, resulting in the nanoporous media on the right side in figure \ref{fig:porous_vs_model}. By choosing a different seed for the initial velocities, it is possible to produce statistically equal samples but with random realizations.
